Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-17 surfaces-write-up/surface-write-up.tex
some minor beamer edits
move background into lemmas, 1pt; section labels

\maketitle
\todo{make title shorter so it doesn't overlap with page numbers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%% Introduction %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
For any Weil $\bq$-divisor $D$ on a rational surface $X$, the graded
\textbf{canonical ring} $R(X, D) := \bigoplus_{d \geq 0} H^0(X, dD)$
has connections to rings of Hilbert and Siegel modular forms. In the
study of these canonical rings, it is natural to describe
presentations. We study the generators and relations of canonical
rings for projective spaces $\bp^m$ and Hirzebruch surfaces $\hirz_m$.
When $D$ is a general $\bq$-divisor on any $\bp^m$ or on any
$\hirz_m$, we give bounds on the generators and relations of
$R(X, D)$. In particular, we give a complete description of the
canonical ring when $X = \bp^m$ and $D$ is an $\bq$-divisor
containing one hypersurface and give stronger bounds when
$D$ is any effective $\bq$-divisor on $\bp^m$.

@@

\section{Background}

@@

\section{Lemmas for Generation and Relations}

\begin{prop}
\label{prop:cone-generation}
Let $n \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$, and let $a_i, b_i \in \bz$ with $1 \leq i \leq n.$ Define
\begin{align*}
	\Sigma = \left \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}a_i = \sum_{i=1}^{n}b_i = 0 \right \}
\end{align*}

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t}
\pdeg_i$. Furthermore, every element $\sigma \in \Sigma$ can be
written in a canonical form $\lambda + \sum_{i = 1}^{t} w_i e_i$ 
with $\lambda = \sum_{i = 1}^{r} s_i e_i$ with $0 \leq s_i < 1$, the
first coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$,
and $w_1, \ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

@@

\section{Canonical Rings on Projective Space}
In this section, we bound the degrees of generators and relations for divisors on $\bp^m$, for all $m \geq 1$. The $\bp^1$ case was covered in \cite{dorney:canonical}.

@@

In this subsection, we restrict attention to the case of effective fractional divisors $D \in \bq \otimes \di \bp^m$.

@@

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
m).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  
Then, $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$. \todo{Talk a little about Hirzebruch surfaces before this paragraph; especially, describe what it's sections look like.}

@@

\section{Conclusion}
This paper extends the work of O'Dorney \cite{dorney:canonical} on $\bp^1$ to $\bp^m$ and Hirzebruch Surfaces, bounding the degrees of generation and relations of canonical rings of arbitrary divisors $D$.  The methods for deducing bounds on generators and relations of $\hirz_k$ applied similar methods to that of O'Dorney's for $\bp^1$.

@@

\section{Acknowledgments}
We are grateful to David Zureick-Brown for introducing us to this
subject, for providing invaluable guidance,
and for his mentorship. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.

>>>>

\maketitle
\todo{make title shorter so it doesn't overlap with page numbers}

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Introduction %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:intro}
For any Weil $\bq$-divisor $D$ on a rational surface $X$, the graded
\textbf{canonical ring} $R(X, D) := \bigoplus_{d \geq 0} H^0(X, dD)$
has connections to rings of Hilbert and Siegel modular forms. In the
study of these canonical rings, it is natural to describe
presentations. We study the generators and relations of canonical
rings for projective spaces $\bp^m$ and Hirzebruch surfaces $\hirz_m$.
When $D$ is a general $\bq$-divisor on any $\bp^m$ or on any
$\hirz_m$, we give bounds on the generators and relations of
$R(X, D)$. In particular, we give a complete description of the
canonical ring when $X = \bp^m$ and $D$ is an $\bq$-divisor
containing one hypersurface and give stronger bounds when
$D$ is any effective $\bq$-divisor on $\bp^m$.

@@

\section{Lemmas for Generation and Relations}
\label{sec:lemmas}

@@

Now we give a result describing the generators of a cone via its
extremal rays.
\todo{write more transition}

\begin{prop}
\label{prop:cone-generation}
Let $n \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$, and let
$a_i, b_i \in \bz$ with $1 \leq i \leq n.$ Define
\begin{align*}
	\Sigma := \left \{(d, c_1, \ldots, c_n) \in \bz^{n + 1} : c_i \geq -
	d \alpha_i, 1 \leq i \leq n \text{ and } \sum_{i = 1}^{n} a_i =
	\sum_{i	= 1}^{n}b_i = 0 \right \}
\end{align*}

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
i.e.~ $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.

Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t}
\pdeg_i$. Furthermore, every element $\sigma \in \Sigma$ can be
written in a canonical form $\lambda + \sum_{i = 1}^{t} w_i e_i$ 
with $\lambda = \sum_{i = 1}^{r} s_i e_i$ with $0 \leq s_i < 1$, the
first coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$,
and $w_1, \ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

@@

\section{Canonical Rings on Projective Space}
\label{sec:proj}
In this section, we bound the degrees of generators and relations for divisors on $\bp^m$, for all $m \geq 1$. The $\bp^1$ case was covered in \cite{dorney:canonical}.

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We completely
characterize the canonical rings when $D$ consists of one
hypersurface and give strong bounds on generator and relation
degrees for general effective divisors $D$.

@@

\section{Canonical Rings on Hirzebruch surfaces}
\label{sec:hirz}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
m).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  
Then, $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$. \todo{Talk a little about Hirzebruch surfaces before this paragraph; especially, describe what it's sections look like.}

@@

\section{Conclusion}
\label{sec:conc}
This paper extends the work of O'Dorney \cite{dorney:canonical} on $\bp^1$ to $\bp^m$ and Hirzebruch Surfaces, bounding the degrees of generation and relations of canonical rings of arbitrary divisors $D$.  The methods for deducing bounds on generators and relations of $\hirz_k$ applied similar methods to that of O'Dorney's for $\bp^1$.

@@

\section{Acknowledgments}
\label{sec:ack}
We are grateful to David Zureick-Brown for introducing us to this
subject, for providing invaluable guidance,
and for his mentorship. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.
